`flippant` --- An `R` package for the automated analysis of dithionite scramblase

### 2016-06-16 inst/manuscript/manuscript.rmd
Various edits/corrections;

# Abstract
Dithionite scramblase assays are an established means of probing the lipid 
scrambling activity of protein extracts and purified proteins. While the assay
has yielded insight into the scramblase activity in crude membrane preparations,
functional validation of candidate scramblases, stoichiometry of scramblase 
complexes and probe ATP--dependence of flippases, data analysis in its context 
has remained a task involving many manual steps. With the extension package 
`flippant` to `R`, a free software environment for statistical computing and 
graphics, we introduce an integrated solution for the analysis and 
publication--grade graphical presentation of dithionite scramblase assays and 
demonstrate its utility in revisiting an originally manual analysis from the 
publication record, closely reproducing the reported results.

# Introduction
Scramblases are proteins mediating the bidirectional, mass--action driven 
equilibration of lipids between the leaflets of lipid bilayers that constitute 
biological membranes [@waechter_flippases_2013]. Independent of energy 
equivalents such as ATP they facilitate this ``flipping'' by providing a 
mechanism of transition through the hydrophobic environment of membranes to 
polar lipid head groups, thus reducing the massive energetic cost that process 
would carry otherwise [@kornberg_inside-outside_1971]. The underlying molecular 
mechanisms remain largely enigmatic [for a discussion including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified [@menon_specific_2000]. 
None the less, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2--adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, one may speculate that lipid scrambling may, in
fact, be an intrinsic property of G--protein coupled receptors in general 
[@goren_constitutive_2014].

@@

Starting with the identification of baseline and post--reduction fluorescence 
from fluorometer--generated spectra, via the calculation of the underlying 
statistics and through publication--grade representation of the results, data 
analysis in the context of this assay has until now involved an extensive series
of manual steps, at most supported by spreadsheet calculation facilities
[see @menon_opsin_2011;@goren_constitutive_2014 for extensive step--wise discussion].
Based on the `R` free software environment for statistical computing and 
graphics [@r_core_team_r:_2015] and prominently drawing on the power and 
elegance of the `ggplot2` plotting system [@wickham_ggplot2_2009], we here 
provide `flippant`, a package that comprehensively, fast and reproducibly covers
all data analysis and graphing needs from raw fluorescence spectra produced by 
scramblase assays. A simple parametrizing spreadsheet is the sole required user 
input.

@@

```{r noeval_demonstrate_installation, eval = FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite(pkgs = "flippant")
library(flippant)
```

@@

Based on @menon_opsin_2011, @goren_constitutive_2014 and
@ploier_perturbed_submitted and largely keeping with nomenclature used therein,
data is processed as follows. Input is format checked and defaults are injected
for facultative parameters/columns as appropriate. 
Fluorescence spectra are parsed. This includes algorithmic determination of when 
dithionite was added to the sample using peak detection by continuous wavelet
transformation as implemented in the `wmtsa` package [@percival_wavelet_2000;@constantine_wmtsa:_2013]. 
Acquisition time is aligned for all spectra in the data set such that the zero 
time point henceforth corresponds to the time of addition. 
Pre--dithionite--addition baseline fluorescence $F_{\text{baseline}}$ is 
determined for each spectrum by averaging (median) over the ten values preceding
dithionite addition. Post--dithionite--addition minimum fluorescence 
$F_{\text{min}}$ is analogously calculated from the last ten data points before 
(and including) a default 400 s or the time point of measurement supplied by the
user in the parameter spreadsheet. $F_{\text{min}}$ is volume--corrected based on
supplied reaction volumes with and without dithionite and for each spectrum/data
point a fluorescence reduction is calculated as follows:
$$
y_{\text{raw}} = 1 - \frac{F_{\text{min}}}{F_{\text{baseline}}}
$$
A relative fluorescence reduction $y$ is calculated scaled to a (required) liposomes--only/no--protein control and the probability $P_{\geq 1}$ for a 
liposome to contain at least one scramblase molecules is calculated using
$$
P_{\geq 1} = \frac{(y - y_0)}{(y_{\text{max}} - y_0)}
$$
where $y$ is the relative fluorescence reduction for the proteoliposome 
measurement and $y_0$ for a measurement of protein--free liposomes. 
$y_{\text{max}}$ refers to either the maximal $y$ in the experiment or is 
derived from a monoexponential fit to the data (default) as a precaution for the 
case where the protein/phospholipid titration did not reach the plateau of the 
saturation curve. The behavior is adjustable using the `scale_to` parameter of 
the corresponding functions (see below). Next a monoexponential curve is fitted
to the relationship of $P_{\geq 1}$ to the protein to phospholipid ratio (PPR; 
(mg/mmol)), providing the scramblase--characterizing fit factor $\alpha$. By
default and following @ploier_perturbed_submitted, PPR is scaled by a factor of
0.65 to account for a fraction of the vesicle pool used that is refractory to
scramblase reconstitution. This behavior can be modified using the 
`ppr_scale_factor` parameter and avoided alltogether setting it to `NULL`.

@@

Figure 4C/D in @ploier_perturbed_submitted. The following work flow in a first 
`flippant`s capabilities are demonstrated here by reanalyzing data underlying 
Figure 4C--F in @ploier_perturbed_submitted. The following work flow in a first 
step plots the corresponding spectral traces (Figure 1). Note that algorithmic
alignment to the time of dithionite addition is employed (default) as well as 
custom trimming of the time axis to account for divergent acquisition 
times between experiments. PPR scaling is used with the default factor of 0.65.

@@

The call results in the protein to phospholipid ratio (PPR) plot shown in
Figure 2A. Using the customization facilities provided by `ggplot2` 
[@wickham_ggplot2_2009] as a plotting system the plot's x--axis is trimmed to
avoid visual compression of all data by a single data point from the wild type 
rhodopsin series in experiment 1, where $\text{PPR}=`r max_PPR`$. Dashed lines
are added automatically as graphical representations of the fit constant 
$\alpha$.

>>>>

# Abstract
Dithionite scramblase assays are an established means of probing the lipid 
scrambling activity of protein extracts and purified proteins. While the assay
has yielded insight into the scramblase activity in crude membrane preparations,
functional validation of candidate scramblases, stoichiometry of scramblase 
complexes as well as ATP--dependence of flippases, data analysis in its context 
has remained a task involving many manual steps. With the extension package 
`flippant` to `R`, a free software environment for statistical computing and 
graphics, we introduce an integrated solution for the analysis and 
publication--grade graphical presentation of dithionite scramblase assays and 
demonstrate its utility in revisiting an originally manual analysis from the 
publication record, closely reproducing the reported results.

# Introduction
Scramblases are proteins mediating the bidirectional, mass--action driven 
equilibration of lipids between the leaflets of lipid bilayers that constitute 
biological membranes [@waechter_flippases_2013]. Independent of energy 
equivalents such as ATP they facilitate this ``flipping'' by providing a 
mechanism of transition through the hydrophobic environment of membranes to 
polar lipid head groups, thus reducing the massive energetic cost that process 
would carry otherwise [@kornberg_inside-outside_1971]. The underlying molecular 
mechanisms remain largely enigmatic [for a discussion including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified [@menon_specific_2000]. 
None the less, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta~2$--adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, one may speculate that lipid scrambling may, in
fact, be an intrinsic property of G--protein coupled receptors in general 
[@goren_constitutive_2014].

@@

Starting with the identification of baseline and post--reduction fluorescence 
from fluorometer--generated spectra, via the calculation of the underlying 
statistics and through publication--grade representation of the results, data 
analysis in the context of this assay has until now involved an extensive series
of manual steps, at most supported by spreadsheet calculation facilities
[see @menon_opsin_2011;@goren_constitutive_2014;@ploier_perturbed_submitted for
extensive step--wise discussion].
Based on the `R` free software environment for statistical computing and 
graphics [@r_core_team_r:_2015] and prominently drawing on the power and 
elegance of the `ggplot2` plotting system [@wickham_ggplot2_2009], we here 
provide `flippant`, a package that comprehensively, fast and reproducibly covers
all data analysis and graphing needs from raw fluorescence spectra produced by 
scramblase assays. A simple parametrizing spreadsheet is the sole required user 
input.

@@

```{r noeval_demonstrate_installation, eval = FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite(pkgs = "flippant")
library("flippant")
```

@@

Based on @menon_opsin_2011, @goren_constitutive_2014 as well as
@ploier_perturbed_submitted and largely keeping with nomenclature used therein,
data is processed as follows. Input is format checked and defaults are injected
for facultative parameters/columns as appropriate. 
Fluorescence spectra are parsed. This includes algorithmic determination of when 
dithionite was added to the sample using peak detection by continuous wavelet
transformation as implemented in the `wmtsa` package [@percival_wavelet_2000;@constantine_wmtsa:_2013]. 
Acquisition time is aligned for all spectra in the data set such that the zero 
time point henceforth corresponds to the time of addition. 
Pre--dithionite--addition baseline fluorescence $F_{\text{baseline}}$ is 
determined for each spectrum by averaging (median) over the ten values preceding
dithionite addition. Post--dithionite--addition minimum fluorescence 
$F_{\text{min}}$ is analogously calculated from the last ten data points before 
(and including) a default 400 s or the time point of measurement supplied by the
user in the parameter spreadsheet. $F_{\text{min}}$ is volume--corrected based on
supplied reaction volumes with and without dithionite and for each spectrum/data
point a fluorescence reduction is calculated as follows:
$$
y_{\text{raw}} = 1 - \frac{F_{\text{min}}}{F_{\text{baseline}}}
$$
A relative fluorescence reduction $y$ is calculated scaled to a (required) liposomes--only/no--protein control and the probability $P_{\geq 1}$ for a 
liposome to contain at least one scramblase molecules is calculated using
$$
P_{\geq 1} = \frac{(y - y_0)}{(y_{\text{max}} - y_0)}
$$
where $y$ is the relative fluorescence reduction for the proteoliposome 
measurement and $y_0$ for a measurement of protein--free liposomes. 
$y_{\text{max}}$ refers to either the maximal $y$ in the experiment or is 
derived from a monoexponential fit to the data (default) as a precaution for the 
case where the protein/phospholipid titration did not reach the plateau of the 
saturation curve. The behavior is adjustable using the `scale_to` parameter of 
the corresponding functions (see below). Next a monoexponential curve is fitted
to the relationship of $P_{\geq 1}$ to the protein to phospholipid ratio (PPR; 
(mg/mmol)), providing the scramblase--characterizing fit factor $\alpha$. By
default and following @ploier_perturbed_submitted, PPR is scaled by a factor of
0.65 to account for a fraction of the vesicle pool used that is refractory to
scramblase reconstitution. This behavior can be modified using the 
`ppr_scale_factor` parameter and avoided alltogether setting it to `NULL`.

@@

`flippant`s capabilities are demonstrated here by reanalyzing data underlying 
Figure 4C--F in @ploier_perturbed_submitted. The following work flow in a first 
step plots the corresponding spectral traces (Figure 1). Note that algorithmic
alignment to the time of dithionite addition is employed (default) as well as 
custom trimming of the time axis to account for divergent acquisition 
times between experiments. PPR scaling is used with the default factor of 0.65.

@@

The call results in the protein to phospholipid ratio (PPR) plot shown in
Figure 2A. Using the customization facilities provided by `ggplot2` 
[@wickham_ggplot2_2009] as a plotting system the plot's x--axis is trimmed to
avoid visual compression of all data by a single data point from the wild type 
rhodopsin series in experiment 1, where $\text{PPR}=`r max_PPR`$.
